---
title: "Joining external data to the bioacoustic dataset"
author: ""
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F) # by default, print out all of the chunks with commands
```

# Overview {.tabset .tabset-fade}

This page provides an overview on how to read in and join external data with the class bioacoustic dataset.

Please start by logging into [rstudio.pomona.edu](https://rstudio.pomona.edu).

This table below shows the hypotheses for each group:

```{r, echo=F}
library(magrittr)
## Pull in Google sheet schedule
schedDF <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1b6yIsvcT57wPnHy7H143Kk4__WMJ-fz6UGHPNcnt0O8/edit?usp=sharing",sheet = "FinalVersions",col_types="ccc")

schedDF %>% 
  htmlTable::addHtmlTableStyle(col.rgroup = c("#f4cccc","#fce5cd","#fff2cc","#d9ead3","#c9daf8","#c9daf8","#ead1dc"
  )) %>% # get the colors by navigating to the Gsheet, clicking on the cell, then clicking on the color icon --> Custom. Will show hexcode
  htmlTable::htmlTable(rnames=FALSE) # can't figure out for now how to merge cells - whatever
```

# Preparing data

## 1: Load packages

Below, we will start by loading packages that we'll need. **Remember**: if you get an error for loading a package in your workspace, it might be because it isn't installed. In that case, just run this command **once** this semester:

```
install.packages("package") # replace "package" with the name of the package that you need
```

```{r pkgs}
### Load packages
library("ggplot2") # plotting functions
library("dplyr") # data wrangling functions
library("readr") # reading in tables, including ones online
library("mosaic") # shuffle (permute) our data
library("lubridate") # package to handle datetime objects
```

## 2: Read data

Next, we will pull in our data and inspect it.

```{r data}
### Load in dataset
soundDF <- readr::read_tsv("https://github.com/EA30POM/site/raw/main/data/bioacousticAY22-24.tsv") # read in spreadsheet from its URL and store in soundDF

### Look at the first few rows
soundDF

### Load AQI and/or temperature datasets
aqi_df <- readr::read_csv("https://github.com/EA30POM/site/raw/main/data/claremont_aqi_data.csv")
temp_df <- readr::read_csv("https://github.com/EA30POM/site/raw/main/data/claremont_average_temperature.csv")
```
```{r,eval=F}
### Look at the first few rows in a spreadsheet viewer
soundDF %>% head() %>% View()
```

```{r, eval=F}
### Look at the first few rows of the external data
head(aqi_df)
head(temp_df)
```

## 3: Ensure that the date columns are in the correct format.

Now, we are going to ensure that the date columns are correctly formatted for all 3 datasets.

```{r, eval=F}
### Clean up the data using lubridate for date conversion
  # Bioacoustic datasset
soundDF <- soundDF %>%
  mutate(date = lubridate::ymd(date))

  # AQI dataset
aqi_df <- aqi_df %>%
  mutate(date = lubridate::ymd(date)) 

  # Temp dataset
temp_df <- temp_df %>%
  mutate(date = lubridate::ymd(date))
```

## 4: Joining on the data

Below, I'm going to show you how you can merge the data based on the dates in the data. It is **up to you** to modify this code to join on your external dataset with the class `soundDF` bioacoustic dataset!

```{r, eval=F}
### Merge AQI and temperature data by date
claremont_data <- temp_df %>%
  left_join(aqi_df, by = "date") # Join on the 'date' column

### Check the merged dataset
head(claremont_data)
View(claremont_data) # view in spreadsheet format
```

## 5: Code in its entirety

```{r, eval=F}
### Load packages
library("ggplot2") # plotting functions
library("dplyr") # data wrangling functions
library("readr") # reading in tables, including ones online
library("mosaic") # shuffle (permute) our data
library("lubridate") # package to handle datetime objects

### Load in dataset
soundDF <- readr::read_tsv("https://github.com/EA30POM/site/raw/main/data/bioacousticAY22-24.tsv") # read in spreadsheet from its URL and store in soundDF

### Look at the first few rows
soundDF

### Load AQI and/or temperature datasets
aqi_df <- readr::read_csv("https://github.com/EA30POM/site/raw/main/data/claremont_aqi_data.csv")
temp_df <- readr::read_csv("https://github.com/EA30POM/site/raw/main/data/claremont_average_temperature.csv")

### Look at the first few rows in a spreadsheet viewer
soundDF %>% head() %>% View()

### Look at the first few rows of the external data
head(aqi_df)
head(temp_df)

### Clean up the data using lubridate for date conversion
  # Bioacoustic datasset
soundDF <- soundDF %>%
  mutate(date = lubridate::ymd(date))

  # AQI dataset
aqi_df <- aqi_df %>%
  mutate(date = lubridate::ymd(date)) 

  # Temp dataset
temp_df <- temp_df %>%
  mutate(date = lubridate::ymd(date))

### Merge AQI and temperature data by date
claremont_data <- temp_df %>%
  left_join(aqi_df, by = "date") # Join on the 'date' column

### Check the merged dataset
head(claremont_data)
View(claremont_data) # view in spreadsheet format
```

